---
title: "Larsson2019"
author: 
  - name: Wancen Mu, Hirak Sarkar, Avi Srivastava, Kwangbom Choi, 
          Rob Patro, Michael I. Love
date: "Feb 1, 2021"
abstract: |
  Airpart identifies sets of genes displaying differential
  cell-type-specific allelic imbalance across cell types
  or states, utilizing single-cell allelic counts. It makes use of a
  generalized fused lasso with binomial observations of allelic
  counts to partition cell types by their allelic
  imbalance. Alternatively, a nonparametric method for partitioning
  cell types is offered. The package includes a number of
  visualizations and quality control functions for examining single
  cell allelic imbalance datasets.
output: 
  html_document:
    toc: true
    toc_float: true 
    theme: united
    highlight: tango
vignette: |
  %\VignetteIndexEntry{Practice with Larsson2019 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## Load required libraries
suppressMessages(suppressWarnings(library(data.table)))
suppressMessages(suppressWarnings(library(plyranges)))
suppressMessages(suppressWarnings(library(tidyverse)))
library(airpart)
```

# Load data

First, we load the preprocessed [Larsson 2019](https://www.nature.com/articles/s41586-018-0836-1) data. It includes one allelic counts(`c57`) and the other allelic counts(`cast`) which was derived after removing some low quality cells indicated in the paper. `celltype` is a character vector of all cell types.

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
load("../data/larsson.rda")
cell_meta_unique <- c("mESC","G1","S","G2M")
order_of_development <- order(match(celltype$x,cell_meta_unique))
assay.list<-list(a1=c57[,order_of_development],a2=cast[,order_of_development])
coldata<-data.frame(x=factor(celltype$x[order_of_development],levels=cell_meta_unique))
sce<-SingleCellExperiment(assays=assay.list, colData=coldata)
```

# Create Ratio matrix

Here we add the pseudo count for gene clustering and visualization.
```{r}
sce <- preprocess(sce)
```

Take a look at highest expressed genes
```{r}
counts(sce)[order(rowSums(counts(sce)),decreasing = T)[1:10],sample(dim(sce)[2],5)]
```
# Quality control step
### QC on cells
```{r}
cellQCmetrics<-cellQC(sce,mad_detected=3,mad_spikegenes = 4,spike = "Ercc")
cellQCmetrics
```

Now define cell filtering automatically or users can manually filter out based on `sum`,`detected` and `spikePercent`. 
```{r}
keep_cell <- (
  cellQCmetrics$filter_sum & # sufficient features (genes)
  cellQCmetrics$filter_detected & # sufficient molecules counted
  cellQCmetrics$filter_spike  # sufficient features expressed compared to spike genes, high quality cells
) 
table(keep_cell)
sce<-sce[,keep_cell]
```

### QC on genes
Note that we require genes expressed in at least **25%** cells within each cell types. 
```{r}
featureQCmetric<-featureQC(sce,spike = "Ercc",sd = 0.025)
keep_feature<-(featureQCmetric$filter_celltype &
                 featureQCmetric$filter_sd &
                 featureQCmetric$filter_spike)
table(keep_feature)
sce<-sce[keep_feature,]
```
# Gene clustering

We provide two methods for gene clustering. 

### 1. Gaussian Mixture modeling

Gaussian mixture modeling is the default method for gene
clustering. The scatter plot is shown based on top 2 PCs of the
smoothed allelic ratio data. The argument `plot=FALSE` can be used to
avoid showing the plot. We have found gene clusters more spread out, more extreme allelic
imbalance they have.
```{r}
sce<-geneCluster(sce, G=c(20,24,28))
metadata(sce)$geneCluster
```
#### Exclude chromosome X genes 

Sex typeing allelic imbalance are excluded because we want to focus on genetically-driven AI.
And `GMM` can actually successfully detected chr x and some imprinted genes in one gene cluster as below figure shows there are monoallelic pattern among fibroblast cells but in a coordinate way. So after removing all the genes on chromosome X, we redo the gene cluster step.
```{r}
makeHeatmap(sce,genecluster=26, show_row_names = TRUE)
```

```{r,eval=FALSE}
suppressMessages(suppressWarnings(library(GenomicFeatures)))
suppressMessages(suppressWarnings(library("org.Mm.eg.db")))
txdb <- makeTxDbFromGFF("ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M1/gencode.vM1.annotation.gtf.gz")
g <- genes(txdb)
g$sym <- mapIds(org.Mm.eg.db, sub("\\..*","",names(g)), "SYMBOL", "ENSEMBL")
g <- g[!is.na(g$sym)]
```

```{r}
load("../data/mm9_genome.rda")
gene_chrx<- unname(g$sym[which(seqnames(g)=="chrX")])
sce <- sce[-which(rownames(sce)%in%gene_chrx),]
```

```{r}
sce<-geneCluster(sce, G=5)
metadata(sce)$geneCluster
```
### 2. Hierarchical clustering 

Hierarchical clustering is an alternative way for clustering. Although in this dataset, it has similar number of clusters. We have found Gaussian mixture model approach (more similar genes within cluster, based on visual inspection in PC space and of allelic ratio heatmaps)
```{r}
sce.hc<-geneCluster(sce,method="hierarchical")
metadata(sce.hc)$geneCluster
```

#### Generating null data and shows number of gene cluster is 1

To show that above gene clusters are not false positive, we simulated null allelic ratio
0.5 to all genes and it turns out only has one gene cluster.
```{r}
cts <- counts(sce)
a1 <- matrix(emdbook::rbetabinom(dim(cts)[1]*dim(cts)[2],
                             prob = 0.5,
                             size = cts,
                             theta = 20), ncol = dim(cts)[2])
a2 <- cts - a1
assay.list<-list(a1=a1,a2=a2)
sce.null<-SingleCellExperiment(assays=assay.list, colData=colData(sce))
sce.null <- preprocess(sce.null)
# jpeg(file="C:/Users/UNC Support/OneDrive - University of North Carolina at Chapel Hill/Lab/GRA/plot/cluster_null.jpeg",width = 8, height = 7,units = "in",res=450)
sce.null<-geneCluster(sce.null, G=c(1:4))
# dev.off()
metadata(sce.null)$geneCluster
```
# Running airpart

### Simple summary table of alleic ratio

Quickly look at the weighted mean of allelic ratio for each gene cluster. Identify the interested gene cluster. The following step is a complement of the QC on genes step.  We
recommend users only run *airpart* when the largest ordered allelic
ratio difference > 0.05 for speed concerns. We find that the allelic
ratio of most of the gene clusters in such cases (small absolute
allelic ratio differences) won't provide enough evidence to detect
differential allelic imbalance.
```{r}
summary<-summaryAllelicRatio(sce)
sapply(1:length(summary), function(i) {
  inst <- summary[[i]]
  inst_order <- inst[order(inst$weighted.mean), ]
  max(diff(inst_order$weighted.mean)) > 0.05
})
```
Here we use gene cluster 3 as an example.

### Experiment-wide beta-binomial over-dispersion

We recommend examining the experiment-wide beta-binomial
over-dispersion, which helps to inform whether to use a binomial
likelihood or a nonparametric approach to partitioning the cell types
by allelic imbalance.

For gene cluster 3, we use fussed lasso method to infer partition.
```{r}
estDisp(sce,genecluster = 3)
```

### Modeling using fused lasso with binomial likelihood

*airpart* offers a method for partitioning cell types using the
generalized fused lasso with binomial likelihood, as implemented in
the *smurf* package. Cell types are merged based on their similarity
of allelic ratios, accounting for excess variability on the ratio from
low counts. The penalization is determined using deviance on held-out
data, with a 1 SE cross-validation rule for favoring smaller models
(more fused cell types).

```{r}
sce_sub <- fusedLasso(sce,
  model = "binomial",
  genecluster = 3, ncores = 4
)
```

Use table and heatmap to visualize the result.
```{r, results="asis"}
# knitr::kable(metadata(sce_sub)$partition, row.names = FALSE)
# jpeg(file="C:/Users/UNC Support/OneDrive - University of North Carolina at Chapel Hill/Lab/GRA/plot/heatmap_larsson.jpeg",width = 8, height = 7,units = "in",res=450)
makeHeatmap(sce_sub, order_by_group = FALSE,show_row_names = F)
# dev.off()
```
Here `ncores` is the number of CPU used for parallelization. Note, the computation time increases approximate linearly as number of genes increases. If you run `niter>1`, please use a `consensusPart` to derive the final partition.


### Calculating allelic ratio estimate via beta-binomial

After *airpart* determines a partition of cell types either by the fused
lasso with binomial likelihood or the nonparametric approach described
above, it give fused lasso estimate or weighted mean as prior Cauchy mean to estimate the allelic ratio per gene and per cell type (and credible intervals) via beta-binomial loglikelihood, using functionality from the *apeglm* package.

Note that the estimates and credible intervals are not equal for genes, because in this step we re-estimate
the conditional cell type means per partition group and accounting for each gene overdispersion. The dispersion parameter is using Efron-like formula shrinkage.


```{r}
sce_sub <- allelicRatio(sce_sub)
svalue <- extractResult(sce_sub,estimates = "svalue")
lower <- extractResult(sce_sub,"lower")
upper <- extractResult(sce_sub,"upper")
indexci<- which(lower[,2]<0.5 & upper[,2]>0.5)
indexs<- which(as.numeric(svalue[,2])>0.05)
```
A violin plot with posterior mean allelic ratios on the y-axis:
note that we preprocess the allelic ration according to strain, so the maternal allelic ratio for fibroblast is 1-estimates

```{r, echo=FALSE, message=FALSE}
suppressMessages(suppressWarnings(library(forestplot)))
suppressMessages(suppressWarnings(library(RColorBrewer)))
makeViolin <- function(sce, xlab = "cell type") {
  ar <- rowData(sce)[, c(grep("ar", colnames(rowData(sce)), value = TRUE))] %>%
    `colnames<-`(levels(sce$x))
  ar[,2:4]<-1-ar[,2:4] %>% as.matrix()
  dat <- data.frame(
    ratio = as.vector(unlist(ar)),
    x = factor(rep(levels(sce$x), each = length(sce))),
    part = factor(rep(metadata(sce)$partition$part, each = length(sce)))
  )
  sample_size <- data.frame(table(sce$x)) %>% `colnames<-`(c("x", "n"))
  dat <- dat %>%
    left_join(sample_size)
  dat$myaxis <- paste0(dat$x, "\n", "n=", dat$n) %>% as.factor()
  dat$myaxis <- factor(dat$myaxis, levels = unique(paste0(dat$x, "\n", "n=", dat$n)))
  p <- ggplot(dat, aes(x = .data$myaxis, y = .data$ratio, fill = .data$part)) +
    geom_violin(color = "#A4A4A4", size = 1.2, alpha = .7) +
    geom_boxplot(width = 0.1, color = "black") +
    theme_minimal() +ylim(0,1)+
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.border = element_rect(fill = "transparent")
    ) +
    scale_fill_brewer(palette = "Set2") +
    labs(x = xlab, y = "maternal allelic ratio")
  p
}
makeForest <- function(sce, genepoi, ctpoi = seq_len(nlevels(sce$x)), showtext = FALSE, xticks, boxsize = .25,
                       xlab = "Allelic Ratio", col,
                       grid = structure(seq(0.1, 0.9, 0.1),
                         gp = gpar(lty = 2, col = "#CCCCFF")
                       ), ...) {
    qual_col_pals <- brewer.pal.info[brewer.pal.info$category == "qual", ]
    col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
  if (missing(xticks)) {
    xticks <- seq(from = 0, to = 1, by = 0.1)
    xtlab <- rep(c(TRUE, FALSE, TRUE, FALSE, FALSE), length.out = length(xticks))
    attr(xticks, "labels") <- xtlab
  }
  if (missing(col)) {
    col <- fpColors(box = col_vector[seq_len(length(ctpoi))])
  }
  ar <- rowData(sce)[, c(grep("svalue", colnames(rowData(sce)), value = TRUE))] %>% `colnames<-`(levels(sce$x))
  if (showtext) {
    forest_text <- data.frame(`Gene` = rownames(sce), ar[, ctpoi])
  } else {
    forest_text <- data.frame(Gene = rownames(sce)) %>% `rownames<-`(rownames(ar))
  }
  if (missing(genepoi)) {
    smin <- apply(apply(ar, 2, as.numeric), 1, min)
    genepoi <- head(order(smin), min(40, nrow(sce)))
  }
  forest_text <- rbind(colnames(forest_text), forest_text[genepoi, ] %>% as.data.frame())
  message("svalue shown in columns per cell type")
  mean = rbind(rep(NA, length(ctpoi)), rowData(sce)[genepoi, c(grep("ar", colnames(rowData(sce)), value = TRUE))[ctpoi]] %>% as.matrix())
  mean[,2:4] = 1 - mean[,2:4]
  forestplot::forestplot(forest_text,
    is.summary = c(TRUE, rep(FALSE, nrow(forest_text) - 1)),
    hrzl_lines = list("2" = gpar(lty = 2)),
    line.margin = .1, # We need to add this to avoid crowding
    legend = levels(sce$x)[ctpoi], grid = grid,
    txt_gp = fpTxtGp(
      label = list(gpar(cex = 0.8), gpar(cex = 0.8, col = "#703C3C")),
      ticks = gpar(cex = 0.9), xlab = gpar(cex = 0.9)
    ),
    boxsize = boxsize, graphwidth = unit(15, "cm"),
    mean = mean,
    lower = rbind(rep(NA, length(ctpoi)), rowData(sce)[genepoi, c(grep("lower", colnames(rowData(sce)), value = TRUE))[ctpoi]] %>% as.matrix()),
    upper = rbind(rep(NA, length(ctpoi)), rowData(sce)[genepoi, c(grep("upper", colnames(rowData(sce)), value = TRUE))[ctpoi]] %>% as.matrix()),
    clip = c(0.01, 1), xticks = xticks, ref = 0.5,
    col = col, xlab = xlab
  )
}
```

```{r}
makeViolin(sce_sub)
```

```{r, warning=FALSE, fig.width=12}
# jpeg(file="C:/Users/wancenmu/OneDrive - University of North Carolina at Chapel Hill/Lab/Github/airpartpaper/plot/forest.jpeg",width = 8, height = 6,units = "in",res=450)
makeForest(sce_sub,showtext = FALSE)
# dev.off()
```

# GO term analysis

Through GO term evaluation, we can find indeed `response to leukemia inhibitory factor` GO terms(GO:1990823) has the most significant p value = 0.0000283.
```{r, results="asis"}
suppressMessages(suppressWarnings(library(goseq)))
i=3
feat<-which(rowData(sce)$cluster==i)
genes <- rep(0,nrow(sce))
names(genes) <- rownames(sce)
genes[feat] <- 1
table(genes)
suppressMessages(pwf <- nullp(genes, "mm9", "geneSymbol"))
suppressMessages(gores <- goseq(pwf, "mm9", "geneSymbol")) 
p_adjust <- p.adjust(gores$over_represented_pvalue,method = "fdr")
gores$`p.adjust`<-p_adjust
knitr::kable(head(gores[gores$ontology=="BP" & gores$numInCat>15,], 15))
```

# Enrichment analysis
To access whether this gene cluster has significant differential allelic imbalance pattern other than null cluster and is correlated with enhancer activity, enrichment analysis was performed between C57 allele h3k27ac peaks and the DAI gene cluster. 

### load non-allelic chip-seq data 
```{r}
mESC<-read_narrowpeaks("../data/mESC.narrowPeak",genome_info = "mm9")
colnames(mcols(mESC))<-c("name","score","fold_enrichment","-log(pvalue)","-log(FDR)/log(10)","summit")
mESC_filter<-mESC[which(mcols(mESC)$fold_enrichment>15)]

mef<-read_narrowpeaks("../data/mef.narrowPeak",genome_info = "mm9")
colnames(mcols(mef))<-c("name","score","fold_enrichment","-log(pvalue)","-log(FDR)/log(10)","summit")
mef_filter<-mef[which(mcols(mef)$fold_enrichment>15)]

## Map IDs
suppressMessages(suppressWarnings(library(GenomicFeatures)))
suppressMessages(suppressWarnings(library("org.Mm.eg.db")))
txdb <- makeTxDbFromGFF("ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M1/gencode.vM1.annotation.gtf.gz")
g <- genes(txdb)
g$sym <- mapIds(org.Mm.eg.db, sub("\\..*","",names(g)), "SYMBOL", "ENSEMBL")
g <- g[!is.na(g$sym)]

## Define gene promoters
prom <- promoters(g, upstream = 2000, downstream = 200)
```

### Define helper function
```{r}
## Get allelic imbalance genes
get_overlap<-function(prom,feat,mESC,mef){
  poi<-which(mcols(prom)$sym %in% feat)
  g2<-prom[poi]
  
  mESC_ov<-join_overlap_inner(g2, mESC)
  mef_ov<-join_overlap_inner(g2, mef)
  
 
  mESC_ov2 <- mESC_ov %>%
    group_by(sym) %>%
    summarize(
      fold_enrichment=max(fold_enrichment),
      peak_count=sum(!is.na(fold_enrichment)))
  
  mef_ov2 <- mef_ov %>%
    group_by(sym) %>%
    summarize(
      fold_enrichment=max(fold_enrichment),
      peak_count=sum(!is.na(fold_enrichment)))
  
  enrich<-merge(mESC_ov2,mef_ov2,by="sym",all=T)
  return(enrich)
}
```

### Define Gene set of interest and null sets
```{r}
i=3
feat<-rownames(sce)[which(rowData(sce)$cluster==i)]

## null set
j=1
feat_other<-rownames(sce)[which(rowData(sce)$cluster==j)]

```

Fisher-exact test and permutation test were both used to test the assumption that stem cell enrichment versus fibroblast enrichment is  correlated  with  gene  cluster. 

### Fisher-exact test for countoverlap
```{r}
# Calculate overlap counts
overlaptxb_prom<-get_overlap(prom,feat,mESC_filter,mef_filter)
overlaptxb_prom_total<- c(sum(overlaptxb_prom$peak_count.x,na.rm = T),sum(overlaptxb_prom$peak_count.y,na.rm = T))
overlaptxb_other_prom<-get_overlap(prom,feat_other,mESC_filter,mef_filter)

overlaptxb_otherprom_total<- c(sum(overlaptxb_other_prom$peak_count.x,na.rm = T),sum(overlaptxb_other_prom$peak_count.y,na.rm = T))
## calculate pvalue
res<-fisher.test(matrix(c(overlaptxb_prom_total,overlaptxb_otherprom_total),ncol=2,byrow = T),alternative = "g")
res$p.value

```

### permutation test
Random  sampled  the  same  number  of genes in the interested gene cluster from the null cluster and the interested gene cluster and calculate odds ratio of stem cell verses fibroblast among the  sampled  genes  and  the  left  genes.  Repetitive  500  times.  Sampling distribution of the odds ratio was roughly following Gaussian distribution. And the odds ratio calculated from original data is far away from the sampling distribution, indicating that this gene cluster are significantly differ with null (p <10−4) and it is consistent with Fisher-exact test results.
```{r, eval=T}
library(pbapply)
or<-pbsapply(1:500,function(i){
  feat1 <- sample(c(feat,feat_other),length(feat))
  overlaptxb_prom<-get_overlap(prom,feat1,mESC_filter,mef_filter)
  overlaptxb_prom_total<- c(sum(overlaptxb_prom$peak_count.x,na.rm = T),sum(overlaptxb_prom$peak_count.y,na.rm = T))
  feat_other1 <- setdiff(c(feat,feat_other),feat1)
  overlaptxb_other_prom<-get_overlap(prom,feat_other1,mESC_filter,mef_filter)
  overlaptxb_otherprom_total<- c(sum(overlaptxb_other_prom$peak_count.x,na.rm = T),sum(overlaptxb_other_prom$peak_count.y,na.rm = T))
  res<-fisher.test(matrix(c(overlaptxb_prom_total,overlaptxb_otherprom_total),ncol=2,byrow =T),alternative = "g")
  return(res$estimate)
})
or <- data.frame(or = or)
p<-ggplot(or, aes(x=or)) + 
  geom_histogram(color="black", fill="lightblue",binwidth=0.05) + 
  # geom_density(alpha=.2, fill="#E69F00") +
  geom_vline(aes(xintercept=1),
            color="lightblue", linetype="dashed",size=1) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.border = element_rect(fill = 'transparent')) +
  labs(x = "Odds ratio",y="Frequency")+
  geom_point(
    data=data.frame(x=2.165635,y=0),
    aes(x,y),
    color="red",
    size=4)
p
```

# Session Info
```{r}
sessionInfo()
```

